---
title: "PA1_Template"
author: "Christopher"
date: "January 21, 2018"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
Loading the Data
```{r}
library(ggplot2)
library(plyr)
activity <- read.csv("activity.csv")
```

Processing the Data
```{r}
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime<- as.POSIXct(activity$date, format="%Y-%m-%d")
```
Pulling data without nas
```{r}
clean <- activity[!is.na(activity$steps),]
```

## What is mean total number of steps taken per day?
Calculate the total number of steps taken per day

Summarizing total steps per date
```{r}
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum, )
colnames(sumTable)<- c("Date", "Steps")
```
Make a histogram of the total number of steps taken each day

Creating the historgram of total steps per day
```{r}
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```
Calculate and report the mean and median of the total number of steps taken per day

Mean and Median of Steps
```{r}
as.integer(mean(sumTable$Steps))
as.integer(median(sumTable$Steps))
```
## What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
library(plyr)
library(ggplot2)
clean <- activity[!is.na(activity$steps),]
intervalTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))
p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
maxSteps <- max(intervalTable$Avg)
intervalTable[intervalTable$Avg==maxSteps,1]
```
The maximum number of steps for a 5-minute interval was 206 steps.

The 5-minute interval which had the maximum number of steps was the 835 interval.

## Imputing missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
nrow(activity[is.na(activity$steps),])
```
The total number of rows with steps = 'NA' is 2304.

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

My strategy for filling in NAs will be to substitute the missing steps with the average 5-minute interval based on the day of the week.
```{r}
avgTable <- ddply(clean, .(interval, day), summarize, Avg = mean(steps))
nadata<- activity[is.na(activity$steps),]
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
```
Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")
mergeData <- rbind(clean, newdata2)
```
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(sumTable2)<- c("Date", "Steps")
as.integer(mean(sumTable2$Steps))
as.integer(median(sumTable2$Steps))
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Black")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Grey", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey") )
```
The new mean of the imputed data is 10821 steps compared to the old mean of 10766 steps. That creates a difference of 55 steps on average per day.

The new median of the imputed data is 11015 steps compared to the old median of 10765 steps. That creates a difference of 250 steps for the median.

However, the overall shape of the distribution has not changed.

## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
```
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
library(lattice) 
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```
Yes, the step activity trends are different based on whether the day occurs on a weekend or not. This may be due to people having an increased opportunity for activity beyond normal work hours for those who work during the week.

